---
title: "Right Ventricular Mortality Analysis"
author: "Samuel Ash"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, error = FALSE, echo=FALSE)
```

# Overview
Kaplan Meier analyes of mortality.

## Load Libraries and Data

## Libraries

```{r load_libs}
library(tidyverse)
library(survival)
library(survminer)
library(cowplot)
library(readxl)
library(naniar)
```

# Data
Abbreviated data from GRW dated July 25, 2018.

```{r load data}
RV_raw = read_excel("../data/RV size by arterial pruning and survival.xlsx")
```

## Note that in this dataset:
1) those who are pruned i.e., those with aBV5<median, have arterial_prune=1, and those that are not pruned (aBV5>median) have arterial_prune=0
2) those that have RV_vol=1 are those with large right ventricles (highest quartile) and those with with RV_vol=0 have small right ventricles

## Data cleaning:
1) remove all rows with NA
2) change vital status and days followed to numeric

```{r data cleaning}
# change . to NA and remove those rows
RV_proc <- RV_raw %>% na_if(".") %>% na.omit()
# change days followed and vital status to numeric
RV_proc$days_followed <- as.numeric(RV_proc$days_followed)
RV_proc$vital_status <- as.numeric(RV_proc$vital_status)
```


# Analysis
##Survival
###Effect of RV size only

```{r rv survival, fig.align='left', fig.height=10, fig.width=10}
KMRV <- survfit(Surv(days_followed, vital_status) ~ RV_vol, data=RV_proc)
surv.km.RV <- ggsurvplot(KMRV, data=RV_proc, risk.table = TRUE, pval = TRUE,
                            conf.int = TRUE, 
                            risk.table.y.text.col = T, risk.table.y.text = TRUE, 
                            legend = c(0.22,0.32), legend.title="Right Ventricular Size",
                            legend.labs=c("Lowest Three Quartiles","Highest Quartile"),
                            risk.table.height = 0.2,
                            palette=c("#E7B800", "#2E9FDF"),
                            title="Survival by Right Ventricular Size",
                            font.legend=c(14), font.title=c(20, "bold"),
                            font.x=c(14), font.y=c(14), font.tickslab=c(14),
                            xlab="Years")
surv.km.RV
KMRV.p = as.matrix(surv_pvalue(KMRV, method = "survdiff", test.for.trend = FALSE, combine = FALSE)[,2])
```

Note that p value for RV volume alone is `r KMRV.p`

###Effect of pruning in those with big right ventricles

```{r pruning survival large rv, fig.align='left', fig.height=10, fig.width=10}
KMPrune.bigRV <- survfit(Surv(days_followed, vital_status) ~ arterial_prune, data=filter(RV_proc,RV_vol==1))
surv.km.prune.bigRV <- ggsurvplot(KMPrune.bigRV, data=filter(RV_proc,RV_vol==1), risk.table = TRUE, 
                                  pval = TRUE, conf.int = TRUE, 
                                  risk.table.y.text.col = T, risk.table.y.text = TRUE, 
                            legend = c(0.15,0.32), legend.title="Pruning",
                            legend.labs=c("No Pruning","Pruning"),
                            risk.table.height = 0.2,
                            palette=c("#E7B800", "#2E9FDF"),
                            title="Survival by Pruning in those with Large Right Ventricles",
                            font.legend=c(14), font.title=c(20, "bold"),
                            font.x=c(14), font.y=c(14), font.tickslab=c(14),
                            xlab="Years")
surv.km.prune.bigRV
KMPrune.bigRV.p = as.matrix(surv_pvalue(KMPrune.bigRV, method = "survdiff", test.for.trend = FALSE, combine = FALSE)[,2])
```
Note that p value for pruning in those with a large RV is `r KMPrune.bigRV.p`

###Effect of right ventricular size in those with and without pruning

```{r rv by pruning survival, fig.align='left', fig.height=10, fig.width=20}
# in those without pruning
KMRV.noprune <- survfit(Surv(days_followed, vital_status) ~ RV_vol, data=filter(RV_proc,arterial_prune==0))
surv.km.RV.noprune <- ggsurvplot(KMRV.noprune, data=filter(RV_proc,arterial_prune==0), 
                                 risk.table = TRUE, pval = TRUE, conf.int = TRUE, 
                                 risk.table.y.text.col = T, risk.table.y.text = TRUE, 
                            legend = c(0.15,0.32), legend.title="Right Ventricular Size",
                            legend.labs=c("Small Right Ventricle","Large Right Ventricle"),
                            risk.table.height = 0.2,
                            palette=c("#E7B800", "#2E9FDF"),
                            title="Those without Arterial Pruning",
                            font.legend=c(14), font.title=c(20, "bold"),
                            font.x=c(14), font.y=c(14), font.tickslab=c(14),
                            xlab="Years")
KMRV.noprune.p = as.matrix(surv_pvalue(KMRV.noprune, method = "survdiff", test.for.trend = FALSE, combine = FALSE)[,2])

# in those with purning
KMRV.prune <- survfit(Surv(days_followed, vital_status) ~ RV_vol, data=filter(RV_proc,arterial_prune==1))
surv.km.RV.prune <- ggsurvplot(KMRV.prune, data=filter(RV_proc,arterial_prune==1), risk.table = TRUE, 
                                  pval = TRUE, conf.int = TRUE, 
                                  risk.table.y.text.col = T, risk.table.y.text = TRUE, 
                            legend = c(0.15,0.32), legend.title="Right Ventricular Size",
                            legend.labs=c("Small Right Ventricle","Large Right Ventricle"),
                            risk.table.height = 0.2,
                            palette=c("#E7B800", "#2E9FDF"),
                            title="Those with Arterial Pruning",
                            font.legend=c(14), font.title=c(20, "bold"),
                            font.x=c(14), font.y=c(14), font.tickslab=c(14),
                            xlab="Years")
KMRV.prune.p = as.matrix(surv_pvalue(KMRV.prune, method = "survdiff", test.for.trend = FALSE, combine = FALSE)[,2])

# plot
prune.plots <- list(surv.km.RV.noprune, surv.km.RV.prune)

# Arrange multiple ggsurvplots and print the output
arrange_ggsurvplots(prune.plots, print = TRUE, ncol = 2, nrow = 1)
```
Note that p value for right ventricular volume in those without purning is `r KMRV.noprune.p` and in those with pruning is `r KMRV.prune.p`